Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-04-02 brms_meeg.tex
minor updates to cache

\abstract{Time-resolved electrophysiological measurements such as those
offered by magneto- or electro-encephalography (M/EEG) provide a unique
window onto neural activity underlying cognitive processes. Typically,
researchers are interested in testing whether and when such measures
differ across conditions and/or groups. The conventional approach
consists in conducting mass-univariate statistics through time followed
by some form of multiplicity correction (e.g., FDR, FWER) or
cluster-based inference. However, while allowing efficient error-rates
control, cluster-based methods have an important downside: they shift
the focus of inference from the timepoint to the cluster level, thus
preventing any conclusion to be made about the precise onset or offset
of effects (e.g., differences across conditions or groups). Here, we
introduce a \emph{model-based} approch for analysing M/EEG timeseriesn
such as ERPs or decoding performance through time, and their differences
across conditions or groups. This approach relies on Bayesian
generalised additive multilevel models and outputs the posterior
probabilility of the effect being above 0 (or above chance) at every
timestep, while naturally taking into account the temporal dependencies
and between-subject variability present in such data. Using both
simulated and actual M/EEG data, we show that the proposed approach
largely outperforms conventional methods in estimating the onset and
offset of M/EEG effects, producing more precise and more reliable
estimates. We provide an R package implementing the approach and
illustrate how to integrate it into M/EEG statistical pipelines in
MNE-Python.}

@@

@@

Description of cluster-based approaches (see
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019})\ldots{}
As pointed by the original authors themselves
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007}), ``there There is
a conflict between this interest in localized effects and our choice for
a global null hypothesis: by controlling the FA {[}false alarm{]} rate
under this global null hypothesis, one cannot quantify the uncertainty
in the spatiotemporal localization of the effect''
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019})\ldots{}
In contrast, the proposed approach, based on Bayesian generalised
additive multilevel models, allows quantifying the probability of
effects at the level of timesteps, voxels, sensors (etc), while
naturally taking into account spatiotemporal dependencies.

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). Splines
coefficients are penalised (usually through the squared of the smooth
functions' second derivative) in a way that can be interpreted as a
prior on the ``wiggliness'' of the function\ldots{}

Terminology: \emph{splines} are functions composed of simpler functions.
These simpler functions are basis functions (e.g., a cubic polynomial)
and the set of basis functions is a \emph{basis}. Each basis function
gets its coefficient and the resultant spline is the sum of these
weighted basis functions.

@@

Instead of averaring, obtain the smooth ERP signal from multilevel
GAM\ldots{} less susceptible to outliers (Meulman et al., 2023)\ldots{}

@@

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms} (i.e., weakly informative priors). We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\). The \texttt{brms} package uses the same syntax as the
\texttt{R} package \texttt{mgcv} v 1.9-1 (\citeproc{ref-mgcv}{Wood,
2017b}) for specifying smooth effects. Figure~\ref{fig-post-prob-test}
shows the predictions of this model together with the raw data.

@@

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0+\epsilon\)
(Figure~\ref{fig-post-prob-test}), with \(\epsilon := 0.05\), which can
be interpreted as the smallest effect size of interest.

@@

\caption{\label{fig-post-prob-test}Posterior probability of the ERP
difference (slope) being above 0 according to the GAM.}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-post-prob-test-1.pdf}

We can also express this as the ratio of posterior probabilities (i.e.,
\(p/(1-p)\)) and visualise the timecourse of this ratio superimposed
with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}).

\caption{\label{fig-post-prob-ratio}Ratio of posterior probability
according to the GAM (on a log10 scale). Timesteps above threshold (10)
are highlighted in green.}

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-post-prob-ratio-1.pdf}

@@

Next we fit a hierarchical/multilevel GAM using summary statistics of
ERPs (mean and SD) at the participant level (similar to what is done in
meta-analysis). Although it is possible to fit a GAMM at the
single-trial level, we present a resource-efficient version of the model
that is fitted directly on the by-participant summary statistics (mean
and SD)\ldots{}

@@

We then computed the difference between the true and estimated
onset/offset of the ERP difference
(\(\text{error}:=|\hat{\theta}-\theta|\)), according to various
\texttt{eps} and \texttt{threshold} values. Remember that the signal is
generated from a truncated Gaussian with an objective onset at 160 ms, a
maximum at 250 ms, and an offset at 342 ms. Figure~\ref{fig-onset-error}
shows that the hierarchical GAM can \emph{exactly} recover the true
onset and offset values, given some reasonable choice of \texttt{eps}
and \texttt{threshold} values (e.g., a threshold of 20).

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.00        13            0.16             0.34           0        0.002
2 0.00        14            0.16             0.34           0        0.002
3 0.00        15            0.16             0.34           0        0.002
4 0.00        16            0.16             0.34           0        0.002
5 0.00        17            0.16             0.34           0        0.002
6 0.01        10            0.16             0.34           0        0.002
\end{verbatim}

@@

\caption{\label{fig-onset-error}Error function of onset (left) and
offset (right) estimation according to various eps and threshold values
(according to the hierarchical GAM). Minimum error values are indicated
by red crosses.}

@@

We compared the ability of the GAMM to correctly estimate the onset and
offset of the ERP difference to widely used methods. First, we conducted
mass-univariate t-tests (thus treating each timestep independently) and
identified the onset and offset of the ERP difference as the first and
last values crossing an arbitrary significance threshold
(\(\alpha = 0.01\)). We then followed the same approach but after
applying different forms of multiplicity correction the the p-values. We
compared two methods that control the false discovery rate (FDR) (i.e.,
BH95, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995}; and
BY01, \citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}), one
method that controls the familywise error rate (FWER) (i.e.,
Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm, 1979}), and two
cluster-based methods (permutation with a single threshold and TFCE,
\citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The BH95, BY01,
and Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.35.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.2.4
(\citeproc{ref-changepoint}{Killick et al., 2022a}), and applied
directly to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}). For visualisation
and interpretability purposes, we converted p-values to s-values, which
can be interpreted as bits of surprising information, assuming a null
effect (\citeproc{ref-greenland2019}{Greenland, 2019})
(Figure~\ref{fig-corrections}).

@@

\caption{\label{fig-corrections}Timecourse of squared t-values and
s-values, with true onset (black dashed line) and onsets identified
using the raw (uncorrected) p-values or the corrected p-values (BH, BY,
Holm).}

@@

Onset/offset estimation methods were assessed using the median absolute
error (MAE) and variance of 10.000 simulated datasets\ldots{}

@@

Next, we assessed the performance of the proposed approach on actual MEG
data (decoding results of Nalborczyk et al., in preparation). We
conducted time-resolved multivariate pattern analysis (MVPA), also known
as decoding\ldots{} As a result, we have a timecourse of decoding
performance (ROC AUC), bounded between 0 and 1, for each participant
(for a total of 32 participants). Now, we want to \emph{test} whether
the group-level average decoding accuracy is above chance (i.e., 0.5) at
each timestep (Figure~\ref{fig-decoding-data}). We fitted a similar GAM
as discussed previously, but we replaced the \(\mathrm{Normal}\)
likelihood function by a \(\mathrm{Beta}\) one to account for the
bounded nature of AUC values (between 0 and 1) (for a tutorial on Beta
regression, see \citeproc{ref-coretta2025}{Coretta \& Bürkner, 2025}).
Note that although we chose a basis dimension of \(k=50\), which seems
appropriate for the present data, this choice should be adapted
according to the properties of the modelled data (e.g., signal-to-noise
ratio, prior low-pass filtering, sampling rate, etc) and should be
assessed by the usual model checking tools (e.g., posterior predictive
checks). We also define a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}), defined as the
chance level plus the standard deviation of the (group-level average)
decoding performance during the baseline period. This ensures
that\ldots{}\footnote{Could an alternative approach be to include a
  predictor for ``baseline vs.~after baseline''?}

@@

We assessed the reliability of the proposed approach using a form of
permutation-based split-half reliability (e.g.,
\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}), which consisted
of the following steps:

\begin{itemize}
\tightlist
\item
  Create many (e.g., 10 or 20) half train/test splits of the data
\item
  For each fold, estimate the onset/offset on both splits using all
  methods
\item
  Then summarise the distribution of onset/offset with the median and
  variance across folds
\end{itemize}

This will allow checking that the proposed approach produces reliable
onset/offset estimates.

@@

\caption{\label{fig-decoding-data}Group-level average decoding
performance (N=32) superimposed with the GAM predictions (in blue) and
the region of practical equivalence (ROPE, in orange) computed from the
baseline period. The blue horizontal line indicates the timesteps at
which the posterior probability ratio is equal to or greater than 20.}

@@

This section is divided in two parts. First, we present the results from
the simulation study, assessing the bias and variance of each method
when applied to simulated data in which the ground truth is known.
Second, we present the results obtained when applying the different
methods to actual MEG data (decoding performance through time),
assessing the reliability of each method and the stability of its
estimates.

@@

Figure~\ref{fig-simulation-mae-variance} shows a summary of the
simulation results, revealing that the proposed approach (\texttt{brms})
has the lowest MAE and variance for both the onset and offset
estimates\ldots{}

@@

\caption{\label{fig-simulation-mae-variance}Median absolute error and
variance of onset and offset estimates for each method.}

@@

\caption{\label{fig-simulation-distribution}Distributions of onset and
offset estimates for each method.}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-simulation-distribution-1.pdf}

@@

Figure~\ref{fig-onset-offset} shows the group-level average decoding
performance through time with onset and offset estimates for each
method. The \texttt{brms\_full} method is similar to the \texttt{brms}
method except that the ROPE is defined on the entire dataset rather than
on the split dataset. Overall, this figure shows that both the
\texttt{Raw\ p-value} and \texttt{FDR\ BH95} methods were extremely
lenient, considering that the decoding performance was above chance
before the onset of the stimulus (false positive) and until the end of
the trial. The \texttt{Change\ point} and \texttt{Cluster\ mass} methods
were the most conservative methods, identifying a time window from
approximately +60ms to +500ms. The \texttt{Holm}, \texttt{TFCE},
\texttt{brms}, and \texttt{brms\_full} methods produced somewhat similar
estimates of onset and offset, from approximately +60ms to
+650ms.\footnote{It should be noted that although each method can
  produce several ``clusters'' of timesteps, we only considered the
  first (onset) and last (offset) timesteps identified by each method to
  compute the error (difference).}

@@

\caption{\label{fig-onset-offset}Group-level average decoding
performance through time with onset and offset estimates for each method
(data from Nalborczyk et al., in preparation).}

@@

Figure~\ref{fig-reliability} shows two properties of each method: i) the
median difference between the onset and offset estimates from each data
split and the onset and offset estimates from the full dataset (x-axis)
and ii) and the variance of its onset and offset estimates across data
splits (y-axis). This figure reveals that the \texttt{brms} \emph{onset
and offset} estimates on each split are the closest to the estimates
from the full dataset on average (0ms difference for the onset estimate
and 5ms difference for the offset estimate). The \texttt{Raw\ p-value}
method has similar performance, but given the aberrant estimates it
produces (cf. Figure~\ref{fig-onset-offset}), the result that it is
consistent between data splits and the full dataset is not convincing on
its own. Overall, the figure reveals that for all other methods, split
datasets produce later onset estimates and earlier offset estimates (as
compared to the estimates from the model fitted on the full dataset).

@@

@@

\bibitem[\citeproctext]{ref-greenland2019}
Greenland, S. (2019). Valid {\emph{P}} -Values Behave Exactly as They
Should: Some Misleading Criticisms of {\emph{P}} -Values and Their
Resolution With {\emph{S}} -Values. \emph{The American Statistician},
\emph{73}(sup1), 106--114.
\url{https://doi.org/10.1080/00031305.2018.1529625}

@@

@@

@@

@@

Now, we want to test whether and when decoding performance is above
chance level (0.5 for a binary decoding task). These two models are
computationally heavier to fit (more observations and 2D smooth
functions)\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting a GAM with two temporal dimensions}
\NormalTok{timegen\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# 2D thin{-}plate spline (tp)}
\NormalTok{    auc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{t2}\NormalTok{(train\_time, test\_time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ timegen\_data,}
    \AttributeTok{family =} \FunctionTok{Beta}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/timegen\_gam\_t2.rds"}
\NormalTok{    )}

\caption{Posterior probability of decoding accuracy being above chance
level (2D GAM).}

\section{Mathematical formulation of the bivariate
GAM}\label{mathematical-formulation-of-the-bivariate-gam}

@@

where \(\left\|\mathbf{s}_i-\mathbf{k}_{m}\right\|\) is the Euclidean
distance between the covariate \(\mathbf{s}_{i}\) and the knot location
\(\mathbf{k}_{m}\).

>>>>

\abstract{Time-resolved electrophysiological measurements such as those
obtained through magneto- or electro-encephalography (M/EEG) offer a
unique window into the neural activity underlying cognitive processes.
Researchers are often interested in determining whether and when these
signals differ across experimental conditions or participant groups. The
conventional approach involves mass-univariate statistical testing
across time, followed by corrections for multiple comparisons (e.g.,
FDR, FWER) or cluster-based inference. While effective for controlling
error rates, cluster-based methods come with a significant limitation:
they shift the focus of inference from individual time points to
clusters, making it difficult to draw precise conclusions about the
onset or offset of observed effects. Here, we introduce a
\emph{model-based} approch for analysing M/EEG timeseries such as
event-related potentials (ERPs) or decoding performance over time. Our
method leverages Bayesian generalised additive multilevel models
(GAMMs), providing posterior probabilities that an effect is above zero
(or above chance) at each time point, while naturally accounting for
temporal dependencies and between-subject variability. Using both
simulated and actual M/EEG datasets, we demonstrate that this approach
substantially outperforms conventional methods in estimating the onset
and offset of neural effects, yielding more precise and reliable
results. We also provide an R package implementing the method and show
how it can be seamlessly integrated into M/EEG analysis pipelines using
MNE-Python.}

@@

Understanding the temporal dynamics of cognitive processes requires
methods that can capture fast-changing neural activity with high
temporal resolution. Magnetoencephalography (MEG) and
electroencephalography (EEG) are two such methods, widely used in
cognitive neuroscience for their ability to track brain activity at the
millisecond scale. These techniques provide rich time series data that
reflect how neural responses unfold in response to stimuli or tasks. A
central goal in many M/EEG studies is to determine whether---and
when---neural responses differ across experimental conditions or
participant groups.

@@

Description of cluster-based approaches (see
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019})\ldots{}

However, as pointed by the original authors themselves
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007}), ``there is a
conflict between this interest in localized effects and our choice for a
global null hypothesis: by controlling the FA {[}false alarm{]} rate
under this global null hypothesis, one cannot quantify the uncertainty
in the spatiotemporal localization of the effect''
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). In
contrast, the proposed approach, based on Bayesian generalised additive
multilevel models, allows quantifying the posterior probability of
effects being above chance at the level of timesteps, voxels, sensors
(etc), while naturally taking into account spatiotemporal dependencies
present in M/EEG timeseries.

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). Splines
coefficients are penalised (usually through the squared of the smooth
functions' second derivative) in a way that can be interpreted, in
Bayesian terms, as a prior on the ``wiggliness'' of the function. In
other words, more complex (wiggly) basis function are penalised. To
clarify the terminology at this stage: \emph{splines} are functions
composed of simpler functions. These simpler functions are basis
functions (e.g., a cubic polynomial) and the set of basis functions is a
\emph{basis}. Each basis function gets its coefficient and the resultant
spline is the sum of these weighted basis functions.

@@

Instead of averaging, obtain the smooth ERP signal from multilevel
GAM\ldots{} less susceptible to outliers
(\citeproc{ref-meulman2023}{Meulman et al., 2023})\ldots{}

@@

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms} (i.e., weakly informative priors). We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\). The \texttt{brms} package uses the same syntax as the
\texttt{R} package \texttt{mgcv} v 1.9-1 (\citeproc{ref-mgcv}{Wood,
2017b}) for specifying smooth effects. Figure~\ref{fig-plot-post-slope}
shows the predictions of this model together with the raw data.

@@

\CommentTok{\# timing the model fitting}
\CommentTok{\# around 600s for 2000 iterations on 4 chains and 4 cores}
\CommentTok{\# around 190s for 2000 iterations on 4 chains and 4 cores + cmdstanr and 01}
\CommentTok{\# around 440s for 2000 iterations on 4 chains and 4 cores + threading(2), cmdstanr and 01}
\CommentTok{\# library(tictoc)}
\CommentTok{\# tic()}

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
    \CommentTok{\# eeg \textasciitilde{} condition + s(time, bs = "cr", k = 20, by = condition),}
    \CommentTok{\# thin{-}plate regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
    \DocumentationTok{\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#}
    \CommentTok{\# speeding up the model fitting \#}
    \DocumentationTok{\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#}
    \CommentTok{\# backend = "cmdstanr",}
    \CommentTok{\# threads = threading(2),}
    \CommentTok{\# stan\_model\_args = list(stanc\_options = list("O1") )}
\NormalTok{    )}

\CommentTok{\# stop timing}
\CommentTok{\# toc()}
\end{Highlighting}
\end{Shaded}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0+\epsilon\)
(Figure~\ref{fig-post-prob-ratio}, left), with \(\epsilon := 0.05\),
which can be interpreted as the smallest effect size of interest. We can
also express this as the ratio of posterior probabilities (i.e.,
\(p/(1-p)\)) and visualise the timecourse of this ratio superimposed
with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}, right).

@@

\caption{\label{fig-post-prob-ratio}Left: Posterior probability of the
ERP difference (slope) being above 0 according to the GAM. Right: Ratio
of posterior probability according to the GAM (on a log10 scale).
Timesteps above threshold (10) are highlighted in green.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-post-prob-ratio-1.pdf}

@@

Next we fit a multilevel version of the GAM. Although it is possible to
fit a GAMM at the single-trial level, we present a resource-efficient
version of the model that is fitted directly on the by-participant
summary statistics (mean and SD, similar to what is done in
meta-analysis\ldots{}

@@

We then computed the difference between the true and estimated
onset/offset of the ERP difference
(\(\text{error}:=|\hat{\theta}-\theta|\)), according to various
\texttt{k} (GAM basis dimension) and \texttt{threshold} values. Remember
that the signal is generated from a truncated Gaussian with an objective
onset at 160 ms, a maximum at 250 ms, and an offset at 342 ms.
Figure~\ref{fig-onset-error} shows that the hierarchical GAM can
\emph{exactly} recover the true onset and offset values, given some
reasonable choice of \texttt{k} and \texttt{threshold} values (e.g., a
threshold of 20). We provide more detailed recommendations on how to set
\texttt{k} in Section~\ref{sec-basis}.

@@

\caption{\label{fig-onset-error}Error function of onset (left) and
offset (right) estimation according to various basis dimension and
threshold values (according to the multilevel GAM).}

@@

To provide a more exhaustive assessment of the performance of the
proposed approach, we compared the ability of the GAMM to correctly
estimate the onset and offset of the ERP difference to widely-used
methods. First, we conducted mass-univariate t-tests (thus treating each
timestep independently) and identified the onset and offset of the ERP
difference as the first and last values crossing an arbitrary
significance threshold (\(\alpha = 0.05\)). We then followed the same
approach but after applying different forms of multiplicity correction
to the p-values. We compared two methods that control the false
discovery rate (FDR) (i.e., BH95, \citeproc{ref-benjamini1995}{Benjamini
\& Hochberg, 1995}; and BY01, \citeproc{ref-benjamini2001}{Benjamini \&
Yekutieli, 2001}), one method that controls the family-wise error rate
(FWER) (i.e., Holm--Bonferroni method, \citeproc{ref-holm1979}{Holm,
1979}), and two cluster-based permutation methods (permutation with a
single cluster-forming threshold and threshold-free cluster enhancement,
TFCE, \citeproc{ref-smith2009}{S. Smith \& Nichols, 2009}). The BH95,
BY01, and Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.42.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.3
(\citeproc{ref-changepoint}{Killick et al., 2022a}), and applied
directly to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}).
Figure~\ref{fig-corrections} illustrates the onsets and offsets
estimated by each method on a single simulated dataset and shows that
all methods over-estimate the true onset and under-estimate the true
offset.

@@

\caption{\label{fig-corrections}Exemplary timecourse of squared t-values
with true onset and offset (vertical black dashed lines) and
onsets/offsets identified using the raw p-values, the corrected p-values
(BH95, BY01, Holm), the cluster-based methods (Cluster-mass, TFCE), or
using the binary segmentation method (Change point).}

@@

The various methods were assessed using the median error and variance of
onset/offset estimates computed on 10.000 simulated datasets\ldots{}

@@

Next, we assessed the performance of the proposed approach on actual MEG
data (decoding results from \citeproc{ref-nalborczyk:inprep}{Nalborczyk
et al., in preparation}). We conducted time-resolved multivariate
pattern analysis (MVPA), also known as decoding\ldots{} As a result, we
have a timecourse of decoding performance (ROC AUC), bounded between 0
and 1, for each participant (for a total of 32 participants). Now, we
want to \emph{test} whether the group-level average decoding accuracy is
above chance (i.e., 0.5) at each timestep
(Figure~\ref{fig-decoding-data}). We fitted a similar GAM as discussed
previously, but we replaced the \(\mathrm{Normal}\) likelihood function
by a \(\mathrm{Beta}\) one to account for the bounded nature of AUC
values (between 0 and 1) (for a tutorial on Beta regression, see
\citeproc{ref-coretta2025}{Coretta \& Bürkner, 2025}). Note that
although we chose a basis dimension of \(k=50\), which seems appropriate
for the present data, this choice should be adapted according to the
properties of the modelled data (e.g., signal-to-noise ratio, prior
low-pass filtering, sampling rate, etc) and should be assessed by the
usual model checking tools (e.g., posterior predictive checks, see also
Section~\ref{sec-basis}). To distinguish signal from noise, we also
defined a region of practical equivalence (ROPE,
\citeproc{ref-kruschke2017}{Kruschke \& Liddell, 2017}), defined as the
chance level plus the standard deviation of the (group-level average)
decoding performance during the baseline period.

@@

We assessed the reliability of the proposed approach using a form of
permutation-based split-half reliability (as for instance in
\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}), which consisted
of the following steps. First, we created 1000 split halves of the data
(i.e., with half the participants in the original data, that is, 16
participants). For each split, we estimated the onset/offset using all
methods described previously. Third, we summarised the distribution of
onset/offset estimates using the median ``error'' (i.e., difference
between the split estimate and the estimate obtained using the full
dataset) and the variance across splits.

@@

\caption{\label{fig-decoding-data}Group-level average decoding
performance (N=32) superimposed with the GAM predictions (in blue) and
the region of practical equivalence (ROPE, in orange) computed from the
baseline period. The blue horizontal markers indicate the timesteps at
which the posterior probability ratio exceeds 20.}

@@

This section is divided in two parts. First, we present the results from
the simulation study, assessing the bias and variance of each method
when applied to simulated data in which the ground truth is known.
Second, we present the results obtained when applying the different
methods to actual MEG data (decoding performance through time),
assessing the reliability of the estimates provided by each method.

@@

Figure~\ref{fig-simulation-mae-variance} shows a summary of the
simulation results, revealing that the proposed approach (\texttt{BGAM})
has the lowest median error and variance for both the onset and offset
estimates.

@@

\caption{\label{fig-simulation-mae-variance}Median error and variance of
onset and offset estimates for each method. Variance is plotted on a
log10 scale for visual purposes.}

@@

@@

Figure~\ref{fig-onset-offset} shows the group-level average decoding
performance through time with onset and offset estimates for each
method. Overall, this figure shows that both the \texttt{Raw\ p-value}
and \texttt{FDR\ BH95} methods were extremely lenient, considering that
the decoding performance was above chance before the onset of the
stimulus (false positive) and until the end of the trial. The
\texttt{Change\ point} and \texttt{Cluster\ mass} methods were the most
conservative methods, identifying a time window from approximately +60ms
to +500ms. The \texttt{Holm}, \texttt{TFCE}, and \texttt{BGAM} methods
produced similar estimates of onset and offset, ranging from
approximately +60ms to +650ms.\footnote{It should be noted that although
  each method can produce several ``clusters'' of ``significant''
  timesteps, we only considered the first (onset) and last (offset)
  timesteps identified by each method to compute the error (difference).}

@@

\caption{\label{fig-onset-offset}Group-level average decoding
performance through time with onset and offset estimates for each method
(data from \citeproc{ref-nalborczyk:inprep}{Nalborczyk et al., in
preparation}).}

@@

Figure~\ref{fig-reliability} shows the median difference between the
onset and offset estimates from each data split and the onset and offset
estimates from the full dataset (x-axis) along with the variance of its
onset and offset estimates across data splits (error bar). This figure
reveals that the \texttt{BGAM} \emph{onset and offset} estimates on each
split are the closest to the estimates from the full dataset on average
(0ms difference for the onset estimate and 5ms difference for the offset
estimate). The \texttt{Raw\ p-value} method has similar performance, but
given the aberrant estimates it produces (cf.
Figure~\ref{fig-onset-offset}), the fact that it is consistent between
data splits and the full dataset is not convincing on its own. The
\texttt{Change\ point} method also has a very good performance (i.e.,
very low difference between split estimates and full estimates), but
produces too short cluster of significant decoding performance (cf.
Figure~\ref{fig-onset-offset}). Overall, the figure reveals that for all
other methods, split datasets produce later onset estimates and earlier
offset estimates (as compared to the estimates from the model fitted on
the full dataset). These results highlight that the results of any
method should be assessed using both i) good asymptotic properties on
simulated data, ii) sensible identified clusters in actual data, and
iii) reliable/stable estimates on actual data.

@@

\section{Acknolwedgements}\label{acknolwedgements}

Centre de Calcul Intensif d'Aix-Marseille is acknowledged for granting
access to its high performance computing resources.

@@

@@

\bibitem[\citeproctext]{ref-nalborczyk:inprep}
Nalborczyk, L., Hauw, F., Torcy, H. de, Dehaene, S., \& Cohen, L. (in
preparation). \emph{Neural and representational dynamics of tickertape
synesthesia}.

@@

\bibitem[\citeproctext]{ref-rasmussen2005}
Rasmussen, C. E., \& Williams, C. K. I. (2005). \emph{Gaussian Processes
for Machine Learning}.
\url{https://doi.org/10.7551/mitpress/3206.001.0001}

@@

\bibitem[\citeproctext]{ref-riutort-mayol_practical_2023}
Riutort-Mayol, G., Bürkner, P.-C., Andersen, M. R., Solin, A., \&
Vehtari, A. (2023). Practical {Hilbert} space approximate {Bayesian
Gaussian} processes for probabilistic programming. \emph{Statistics and
Computing}, \emph{33}(1), 17.
\url{https://doi.org/10.1007/s11222-022-10167-2}

@@

@@

where \(\left\|\mathbf{s}_i-\mathbf{k}_{m}\right\|\) is the Euclidean
distance between the covariate \(\mathbf{s}_{i}\) and the knot location
\(\mathbf{k}_{m}\). We fitted this model using \texttt{brms}\ldots{}
